On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-09-21 Manuscript/manuscript.tex
polishing; comments on p-values and phi

Devon Allen, an alumnus of the University of Oregon in Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field Championships also in Eugene 
in late June, Allen was poised to compete in the 2022 World Track and 
Field Championships in August, which would also be held in Eugene, Oregon.
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 and 0.101 in his preliminary heat and semifinal, respectively.  
His disqualification in the final heat by reacting in 0.099 seconds, just 0.001
seconds faster than the 0.1 threshold set by the International Association of
Athletics Federations (IAAF), was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification
(\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}).
The crowd erupted with jeers
when they realized that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete. %and been 
%spared waiting until the next World Track and Field Championships.
%\eds{He is not competing this year, correct? If yes, we may need to rephrase.}
%\of{Correct}
%\eds{removed because he did not compete this year, so did not actually have
% to wait, after all.} 
%\eds{ If it is true he ``retired'' as a runner to play football, we could 
%instead say something like
%\textit{Had Allen reacted just 0.001 seconds slower he would have been allowed 
%to compete before retiring from running to play football.}}
%\of{I don't think he retired, he was competing in races as recently as this past
%June.}
%\eds{Okay, then let's leave it like this.}

Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit, is very active during important running
events, including during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as \citet{johnson2022data}
and \citet{johnson2022was} from  Robert Johnson, who created and runs 
the LetsRun website.  He argued that there was an error
with the timing equipment at the 2022 World Track and Field Championships and
cited graphs, computed simple statistics such as median reaction times, and 
compared the reaction times at the 2022 World Track and Field Championships to 
other competitions
such as the United States Track and Field Championships and other World Championships.
The comparisons lacked formal statistical analysis, however. 
The alleged 99.95\% chance Allen was wronged 
came from Johnson claiming there is a $(1/13)^3$ chance that the fastest 
reaction times over the past 13 competitions of the men's 100 meter, men's 110 
meter hurdle, and women's 100 meter would all come from the same year 
\citep{johnson2022was}.

Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} examined reaction times for nine male athletes.  The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by IAAF,
\citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Unfortunately, IAAF has not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.
It is unknown how many athletes since the 2009 study could have potentially
benefited from this, but it is probable that there are others besides Allen who
could have. Additionally,
%were disqualified. Additionally,
\citet{komi2009iaaf} recommended that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been shown to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

@@

The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championships to check for any abnormalities, and we approach
this in two different ways.  First we compare reaction times for athletes who 
have competed at additional competitions to the 2022 World Track and Field 
Championship.  By comparing reaction times from the same athlete across multiple
competitions we are able to determine if the 2022 reaction times differed 
significantly from those of other competitions.  Then we model the data, which are
the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with two random effects: 
one effect for the year and a second effect for the heat within each year.
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Our investigation into reaction times is important not only because of its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times, which is the goal of
every sprinter who steps onto the track.
%\eds{Since the World Championships are happening again now, is data from this 
%year already available? If it is available, we would need to justify its omission
%in the current analysis.
%If it is not available, but becomes available while under review, would we 
%reanalyze the data?  Does it seem likely our conclusions would change based on
%what is going on this year?}
%\of{Resolved}

The rest of this paper is organized as follows. Section~\ref{sec:Data} details 
how the two types of data were collected, as the signed-rank test as described
in Section~\ref{sec:rank} uses slightly different data than the generalized 
linear mixed-effect model (GLMM) that is further developed in 
Section~\ref{sec:glmm}. Section~\ref{sec:Results} provides the results of the two
analyses and ends by questioning the reaction time barrier.  Lastly, 
Section~\ref{sec:concludingremarks} discusses the impact and limitations of the
paper.

@@

There are two types of data that are explored throughout this paper.  For the
signed-rank tests described in Section~\ref{sec:rank}, the data comes from 
athletes who competed in the 2022 World Championships and another competition 
(2019 World Championships, 2023 World Championships, or 2022 national 
championships). For the GLMM described in Section~\ref{sec:glmm}, data comes
from every world championships from 1999 to 2023. This investigation started 
shortly after the 2022 World Championships, but as the 2023 data has now become 
available, we are pleased to include it in both of our analyses, and it did not
significantly change our results \citep{WAData}.
\eds{I think you should somehow reference the source of the data here; I saw the
comment below about removing the Appendix and including the sources in the Data 
README file, so assuming the readers will have access to that file, you can 
simply refer to the sources within that file here. I will defer to Prof. Yan's
opinion, however!}
\of{Is this what you meant?}

\subsection{Comparison Data with 2022 World Championships}\label{sec:databeyond}
The motivation for looking beyond World Championships data came from an 
exploratory analysis examining how United States athletes reacted at the 
2022 United States Track and Field Championships.

\subsubsection{2022 National Competitions}\label{sec:datanational}
From June 23-26, 2022, the United States held its Track and Field Championships 
at Hayward Field in Eugene, Oregon to decide who to send to the World 
Championships being held at the same venue in August. Thus, we can establish a 
baseline for the four United States athletes who competed in at least one round 
of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen.   
Every athlete reacted faster in all the World Track and Field Championships 
races compared to the United States Track and Field Championships races. The 
dataset of the four United States athletes is too small to perform any 
meaningful statistical analysis, but the data was expanded to include reaction 
times for 11 more athletes who competed at other national competitions (many of which were 
held from May-July of 2022). This data was difficult to find, as the data is not
centrally located but instead found on different national websites (often 
recorded in the national language without translation). 
%\of{I don't know how to say this}. \eds{is my change correct?} \of{No I was able
%to use a lot even though I couldn't actually read the page. I just tried to find
%110 meter hurdle and then look at a column for reaction times which are always from
%roughly 0.1 to 0.18 seconds.} \eds{is it correct now?} \of{Yes}

@@

While it is theoretically possible that athletes got faster 
between 2019 and 2022 or 2022 and 2023, it is important to 
remember that these are some of the fastest sprinters on the planet, and thus
improving on elite times is not easy.  Once again we repeat a similar procedure
as in Section~\ref{sec:datanational}; we assign each athlete to be a cluster and
within each cluster label times to be from either the 2019, 2022, or 2023 World
Championships. When comparing 2019 and 2022 we look at data from 15
athletes, with two clusters of size two, three clusters of size three, five 
clusters of size four, two clusters of size five, and three clusters of size six.
When comparing 2022 and 2023 we look at data from 22 athletes, with seven 
clusters of size two, three clusters of size three, six clusters of size four,
and six clusters of size five.  
%Although 2022 is presented three times in
%Figure~\ref{fig:RankScatterplots}, the subset of athletes differs each time.
%There may be athletes who competed in 2019 and 2022 and had their national
%championships data available, but they did not compete in 2023, and thus their
%times appear only in the top two graphs. \eds{top and bottom graphs?}
%\eds{also, you will likely need to make Fig 1 smaller as the caption runs off
%the page (even before my edits).}
%\of{It should be smaller now}

\jy{give specifics of your sample size}
\of{Resolved}

\begin{figure}[tbp]
  \centering
  \includegraphics{RankScatterPlots}
  \caption{The top graph shows the reaction times of athletes who competed at 
	least once each at the 2019 and 2022 World Championships. The middle graph 
	shows the reaction times of athletes who competed at least once each at the 
	2022 and 2023 World Championships. The bottom graph shows the reaction times 
	of athletes who competed at least once at their country's
  2022 national championships and the 2022 World Championships.}
	%\eds{check x-axis labels in the bottom plot: last three should be 2022?}
  %\of{I don't think so. Its a comparison of 2022 national vs international so
  %I abbreviated it as Nat. and Inter. is that not ok?}\eds{it's good now}
  \label{fig:RankScatterplots}
\end{figure}

Figure~\ref{fig:RankScatterplots} shows the data that will be analyzed
with the rank-based methods (described in Section~\ref{sec:rank} with results in 
Section~\ref{subsec:Results_Rank}). In each of the three graphs, the reaction 
times for athletes who competed at the 2022 World Championships 
\eds{World Championships?}
\of{Resolved} are compared to their reaction
times at another at which they competed.  
It is worth noting that each instance of the 2022 data is different, because
those who competed at both the 2019 and 2022 World Championships will be
different athletes than those who competed at both the 2022 and 2023 World
Championships. This data was only for athletes who competed
in at least one race at each championship, however it is possible for athletes
to have competed in one race in 2019 and two races in 2022. 
%\eds{did we not already cover the last two sentences in the previous paragraph, 
%or are you trying to say something different here?} \of{yes you are correct. I
%commented out the other one because I think this one flows better.}  
The 2022 World
Championship Heats and Semifinals both had on average lower reaction times than
their respective 2019 races, and the 2022 finals were substantially faster. It
is also worth noting is that the fastest reaction time in the 2022 World
Championships Finals and Semifinals both came from Devon Allen, but the 
difference between 0.101 and 0.099
seconds was the difference in what caused him to be disqualified.

We also started to perform a similar analysis to compare the times of female 
athletes who competed at the 2019 and 2022 World Track and Field Championships,
however as our results were not significant, we decided to incorporate another
method to investigate the times at the 2022 World Championships and thus we
employ the GLMM, which looks at historical men's data to try to determine
whether Allen's time and disqualification are reasonable. 
\eds{I would suggest removing this paragraph, particularly if you still include
the comments about female reaction times in the Concluding Remarks. 
No analysis has been done yet, nor have the methods been described yet, so it 
does not make sense in my opinion 
to refer to significance or another analysis method here.}
\of{The reason I added it was because Prof Yan recommended switching the
order of glmm and rank and talking about the GLMM partly because the women's
data was not significant. I'm not sure how else to say this or if it needs to
be said at all.  I agree that it doesn't make a ton of sense to talk about the 
results for the women's data so early.  Prof Yan, can you please include your
input? Maybe we keep the rank before glmm but don't bring up women's data until
the end?}

@@

Data was taken from the World Athletics 
and covers the men's 110 meter hurdles from 1999 to 2023.
Although women's data is available, we chose not to include women's data in the
analyses to follow due to the numerous papers citing differences in men and
women's reaction times \citep[e.g.,][]{lipps2011implications, 
babicc2009reaction, panoutsakopoulos2020gender}.
\eds{you should mention the data source earlier, possible in Section 2 
before Section 2.1 if it applies to Section 2.1.1, and also the sources of the 
national data for 2.1.2 (i.e., pointing to the info in Appendix).}
\of{I want to emphasize the differences in the data so I wanted to try and
compartmentalize everything that was different. We chose not to include women's
data for the GLMM because of cited papers and chose not to include women's
data for the signed-rank because the results were not highly significant. So if
I want to talk about that in section 2, I feel like I would have to explicitly
justify the exclusion of the women's data which then goes back to the
conversation above. Also Prof Yan recommended I remove the appendix, so now the
sources are in the Data README.}
\eds{Got it.  See earlier comment in 2.1; I will defer to Prof. Yan.}
For the creation of the GLMM model we focus on the reaction times during the 
semifinal and final heats only, as reaction times from preliminary heats are 
often not as fast as those in later heats \citep[e.g.,][]{zhang2021correlation}.

For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together which increases the sample size. This is 
advantageous because in some years there are very few finals observations, such
as in 2022, when there were only five data points due to two disqualifications 
and one athlete who did not compete.  This is the data that we will reference
throughout the rest of the paper unless otherwise noted. We also consider data 
that excludes 2022 to see how our analysis differs based on the exclusion of one
particular year of interest.

@@

With our small sample size, the asymptotic normal distribution may not be
reliable, so we also use 1000 random permutations to simulate the null 
distribution of the test statistic.  
%\eds{simulate the null distribution of the test statistic?}
%\of{Resolved}
This method is available from the \texttt{clusWilcox.test()} function
with \texttt{method = `ds'} (for \underline{D}atta and \underline{S}atten) and 
\texttt{exact = TRUE} from R package
\texttt{clusrank} \citep{jiang2020wilcoxon}.

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the we chose a gamma mixed-effect model as our 
final model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. Two random effects were found useful: one capturing the venue effect in
each year and the other capturing the heat effect within each year.  The heat
effect is important as it captures the variability in the amount of time
athletes are on the starting blocks before the gun goes off.  Thus, we can try
to reduce this variability by accounting for it in the model through the heat
effect.
Let $Y_{ijk}$ be the reaction time of observation~$k$ in heat~$j$ in year~$i$.
Conditional on a venue effect $v_i$ of year~$i$ and a heat effect
$h_{i/j}$ nested within each year~$i$, $Y_{ijk}$ is modeled by a
gamma distribution. 
%\eds{I was talking to my husband about this. Could the heat effect be justified 
%to capture the variability in the amount of time between the ``set'' and the 
%gun going off (assuming this is different from heat to heat)? E.g., quicker 
%gunshots after the ``set'' could lead to slower reaction times (or the reverse), 
%but this effect would be shared across all athletes in the same heat.}
%\of{Yes I agree with this and updated the language in the text.}
In the classic generalized linear model (GLM) sense,
$Y_{ijk}$ has conditional mean $E(Y_{ijk} | v_i, h_{i/j}) > 0$ and a
dispersion parameter~$\phi > 0$. The conditional mean is characterized by
\[
g\{E(Y_{ijk} | v_i, h_{i/j})\} = \alpha + v_i + h_{i/j},
\]
where $g$ is a known link function,
$v_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $\delta_{i/j}$ 
\eds{$h_{i/j}$ or $\delta_{i/j}$? Table 2 uses $\delta_{i/j}$ (are you reporting
the BLUP for this? If yes, why not also report the BLUP of $v_i$ in Table 2?)}
\of{I think this should say $h_{i/j}$. In fact I am not sure we ever properly 
defined what $\delta_{i/j}$ is. The number in the last column of the table is
what shows up as the residual standard deviation in R}
is normally distributed with mean
zero and variance~$\sigma_h^2$.
Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time in cognitive
psychological research, the logarithm
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

To better understand this model, we can identify the gamma model in terms of the
commonly used shape and scale parameters. For a gamma distribution with
shape~$a$ and scale~$b$, we have mean $\mu = ab$ and variance $\nu = ab^2$. The
variance as a function of the mean~$\mu$ in the GLM sense is
$\text{var}(\mu) = \mu^2 / a$, which implies that the dispersion parameter is
$\phi = 1 / a$. Therefore, for the gamma mixed-effect model, the conditional
gamma distribution of the reaction time $Y_{ijk}$ given the venue
effect~$v_i$ and heat effect~$h_{i/j}$
has shape $a = 1 / \phi$ and scale
$b = g^{-1}(\alpha + v_i + h_{i/j}) \phi$. The marginal
distribution of $Y_{ijk}$ is a scale-mixture of gamma distributions, which can be
easily simulated from once the parameters are estimated. Many
random numbers generated from the fitted mixture distribution can be used to
approximate the probability of observing a reaction time faster than any given
threshold.  We are specifically interested in the probability of a reaction time
being less than 0.1 seconds in order to gauge if that is a reasonable 
disqualification barrier.

@@

Table~\ref{tab:Clusrankresults} shows the results generated from both the 
permutation (exact) and asymptotic rank-based tests for the three comparisons, 
as discussed previously in Section~\ref{sec:rank}.    
%six \eds{three?}
%\of{Theres now 6 results in the table, if we remove the asymptotic results then
%it should be 3}
%tests in R.
%We use a \texttt{clusWilcox.test()} with \texttt{exact = TRUE} and 
%\texttt{B = 1000} indicating that the random permutation size was chosen to be 
%1000, as discussed previously in Section~\ref{sec:rank}.  
%As the sample sizes are small, 65 and 49, we use \texttt{exact = TRUE} as
%that is preferred.  
%In both cases, 
The national versus international comparison
produced lower p values than the 2019 versus 2022 comparison, with the
national vs international permutation test producing a p value of 
$2.2\cdot10^{-16}$, which is the smallest possible value achievable in R.  
\eds{Prof. Yan, is this possible based on 1000 permutations?} This
is a highly significant test result and shows that athletes who competed at the
2022 World Track and Field Championships were on average, significantly faster
at reacting than they were just a few months prior.  For the 2019 versus 2022
same athlete comparison with permutation test, the p value was 
$2.9\cdot10^{-3}$, which is significant at $\alpha = 0.01$.  The result of
comparing 2022 and 2023 was very similar to the above result, as the p value
from the permutation test was $3.0\cdot10^{-3}$.  Taken together, these three
tests help show that athletes competing at the 2022 World Track and Field
Championships had faster reaction times than when they had competed previously.

%\eds{not sure the \texttt{exact = FALSE} results need to be reported here or in
%Table 4 when we say we use \texttt{exact = TRUE} here and in the previous section.}
%\of{I am not going to talk about the asymptotic results but will show them in
%table.I changed the exact=TRUE and exact = FALSE text to say permutation and 
%asymptotic}
%these
%
%\eds{need to comment on 2022 vs 2023 results}
%\of{Resolved}

@@

@@

%One conclusion from the analysis in \citet{zhang2021correlation} was that 
%athletes' reaction times improved from 2011 to 2019.  However this study does 
%not consider data from 1999 to 2009, and over this timespan athletes get slower 
%and then faster.  Figure~\ref{fig:Boxplots} shows that for the men's pooled data,
%the year with the fastest median reaction time before 2022 was 1999.  Thus, this 
%reject's Zhang et al.'s conclusion that athletes are getting faster because they 
%were fast over twenty years ago. One possible explanation for why 2011 had such
%high variation was because of the IAAF rule discussed earlier that increased the
%penalty of a false start to an instant disqualification \citep{iaaf2009falsestart}.
%Athletes may have been overly cautious to false start and may have consciously
%or sub-consciously reacted slower as a result.  Instead, a more appropriate 
%conclusion may be that there is a lot of variability in reaction times and that
%the context of the races matters.

>>>>

Devon Allen, an alumnus of the University of Oregon in Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field Championships also in Eugene 
in late June, Allen was poised to compete in the 2022 World Track and 
Field Championships in August, which would also be held in Eugene, Oregon.
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 and 0.101 in his preliminary heat and semifinal, respectively.  
His disqualification in the final heat by reacting in 0.099 seconds, just 0.001
seconds faster than the 0.1 threshold set by the International Association of
Athletics Federations (IAAF), was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification
(\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}).
The crowd erupted with jeers
when they realized that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete.

Allen's disqualification has been discussed extensively in online communities
such as \url{www.LetsRun.com}, a website that is part message board and part
news. The message board,
which functions similarly to Reddit, is very active during important running
events, including during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as \citet{johnson2022data}
and \citet{johnson2022was} from  Robert Johnson, who created and runs 
the LetsRun website.  He argued that there was an error
with the timing equipment at the 2022 World Track and Field Championships and
cited graphs, computed descriptive statistics such as median reaction times, and 
compared the reaction times at the 2022 World Track and Field Championships to 
other competitions
such as the United States Track and Field Championships and other World Championships.
The comparisons lacked formal statistical analysis, however. 
The alleged 99.95\% chance Allen was wronged 
came from Johnson claiming there is a $(1/13)^3$ chance that the fastest 
reaction times over the past 13 competitions of the men's 100 meter, men's 110 
meter hurdle, and women's 100 meter would all come from the same year 
\citep{johnson2022was}.

Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} examined reaction times for nine male athletes.  The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by IAAF,
\citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Unfortunately, IAAF has not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.
It is unknown how many athletes since the 2009 study could have potentially
benefited from this, but it is probable that there are others besides Allen who
could have. Additionally,
\citet{komi2009iaaf} recommended that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been shown to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

@@

The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championships to check for any abnormalities, and we approach
this in two different ways.  First we compare reaction times for athletes who 
have competed at additional competitions to the 2022 World Track and Field 
Championship.  By comparing reaction times from the same athlete across multiple
competitions we are able to determine if the 2022 reaction times differed 
significantly from those of other competitions.  Then we look at a longer time horizon
and model the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with two random effects: 
one effect for the year and a second effect for the heat nested within each year.
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Our investigation into reaction times is important not only because of its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times, which is the goal of
every sprinter who steps onto the track.

The rest of this paper is organized as follows. Section~\ref{sec:Data} details 
how the two types of data were collected, as the signed-rank test as described
in Section~\ref{sec:rank} uses slightly different data than the generalized 
linear mixed-effect model (GLMM) that is further developed in 
Section~\ref{sec:glmm}. Section~\ref{sec:Results} provides the results of the two
analyses and ends by questioning the reaction time barrier.  Lastly, 
Section~\ref{sec:concludingremarks} discusses the impact and limitations of the
paper. All the data and code of our analysis are available in a supplement.

@@

There are two types of data that are explored throughout this paper.  For the
signed-rank tests described in Section~\ref{sec:rank}, the data comes from 
athletes who competed in the 2022 World Championships and another competition 
(2019 World Championships, 2023 World Championships, or 2022 national 
championships). For the GLMM described in Section~\ref{sec:glmm}, data comes
from every world championships from 1999 to 2023. This investigation started 
shortly after the 2022 World Championships, but as the 2023 data has now become 
available, we are pleased to include it in both of our analyses, and it did not
significantly change our results \citep{WAData}. 
\jy{I added a sentence to the end of intro.}

\subsection{Comparison Data with 2022 World Championships}
\label{sec:databeyond}

The motivation for looking beyond World Championships data came from an 
exploratory analysis examining how United States athletes reacted at the 
2022 United States Track and Field Championships.

\subsubsection{2022 National Competitions}\label{sec:datanational}
From June 23-26, 2022, the United States held its Track and Field Championships 
at Hayward Field in Eugene, Oregon to decide who to send to the World 
Championships being held at the same venue in August. Thus, we can establish a 
baseline for the four United States athletes who competed in at least one round 
of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen.   
Every athlete reacted faster in all the World Track and Field Championships 
races compared to the United States Track and Field Championships races. The 
dataset of the four United States athletes is too small to perform any 
meaningful statistical analysis, but the data was expanded to include reaction 
times for 11 more athletes who competed at other national competitions (many of which were 
held from May-July of 2022). This data was difficult to find, as the data is not
centrally located but instead found on different national websites (often 
recorded in the national language without translation).

@@

While it is theoretically possible that athletes got faster 
between 2019 and 2022 or 2022 and 2023, it is important to 
remember that these are some of the fastest sprinters on the planet, and thus
improving on elite times is not easy.  Once again we repeat a similar procedure
as in Section~\ref{sec:datanational}; we assign each athlete to be a cluster and
within each cluster label times to be from either the 2019, 2022, or 2023 World
Championships. When comparing 2019 and 2022 we look at data from 15
athletes, with two clusters of size two, three clusters of size three, five 
clusters of size four, two clusters of size five, and three clusters of size six.
When comparing 2022 and 2023 we look at data from 22 athletes, with seven 
clusters of size two, three clusters of size three, six clusters of size four,
and six clusters of size five.

\begin{figure}[tbp]
  \centering
  \includegraphics{RankScatterPlots}
  \caption{The top graph shows the reaction times of athletes who competed at 
	least once each at the 2019 and 2022 World Championships. The middle graph 
	shows the reaction times of athletes who competed at least once each at the 
	2022 and 2023 World Championships. The bottom graph shows the reaction times 
	of athletes who competed at least once at their country's
        2022 national championships and the 2022 World Championships.}
  \label{fig:RankScatterplots}
\end{figure}

Figure~\ref{fig:RankScatterplots} shows the data that will be analyzed
with the rank-based methods (described in Section~\ref{sec:rank} with results in 
Section~\ref{subsec:Results_Rank}). In each of the three graphs, the reaction 
times for athletes who competed at the 2022 World Championships
are compared to their reaction times at another at which they competed.  
It is worth noting that each instance of the 2022 data is different, because
those who competed at both the 2019 and 2022 World Championships will be
different athletes than those who competed at both the 2022 and 2023 World
Championships. This data was only for athletes who competed
in at least one race at each championship, however it is possible for athletes
to have competed in one race in 2019 and two races in 2022.
The 2022 World
Championship Heats and Semifinals both had on average lower reaction times than
their respective 2019 races, and the 2022 finals were substantially faster. It
is also worth noting is that the fastest reaction time in the 2022 World
Championships Finals and Semifinals both came from Devon Allen, but the 
difference between 0.101 and 0.099
seconds was the difference in what caused him to be disqualified.

@@

Data was taken from the World Athletics 
and covers the men's 110 meter hurdles from 1999 to 2023.
We focus on the reaction times during the 
semifinal and final heats only, as reaction times from preliminary heats are 
often not as fast as those in later heats \citep[e.g.,][]{zhang2021correlation}. 
For analysis purposes, we will pool reaction times of men's final and 
men's semifinal heats together which increases the sample size. This is 
advantageous because in some years there are very few finals observations, such
as in 2022, when there were only five data points due to two disqualifications 
and one athlete who did not compete.  This is the data that we will reference
throughout the rest of the paper unless otherwise noted. We also consider data 
that excludes 2022 to see how our analysis differs based on the exclusion of one
particular year of interest.

@@

With our small sample size, the asymptotic normal distribution may not be
reliable, so we also use 1000 random permutations to simulate the null 
distribution of the test statistic.  \jy{increase 1000 to a much bigger number,
  say 10 million}
This method is available from the \texttt{clusWilcox.test()} function
with \texttt{method = `ds'} (for \underline{D}atta and \underline{S}atten) and 
\texttt{exact = TRUE} from R package
\texttt{clusrank} \citep{jiang2020wilcoxon}.

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the we chose a gamma mixed-effect model as our 
final model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. Two random effects were found useful: one capturing the venue effect in
each year and the other capturing the heat effect within each year.  The heat
effect is important as it captures the variability in the amount of time
athletes are on the starting blocks before the gun goes off.  Thus, we can try
to reduce this variability by accounting for it in the model through the heat
effect.

Let $Y_{ijk}$ be the reaction time of observation~$k$ in heat~$j$ in year~$i$.
Conditional on a venue effect $v_i$ of year~$i$ and a heat effect
$h_{i/j}$ nested within each year~$i$, $Y_{ijk}$ is modeled by a
gamma distribution. In the classic generalized linear model (GLM) sense,
$Y_{ijk}$ has conditional mean $E(Y_{ijk} | v_i, h_{i/j}) > 0$ and a
dispersion parameter~$\phi > 0$. The conditional mean is characterized by
\[
g\{E(Y_{ijk} | v_i, h_{i/j})\} = \alpha + v_i + h_{i/j},
\]
where $g$ is a known link function,
$v_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $\delta_{i/j}$ 
is normally distributed with mean
zero and variance~$\sigma_h^2$.
Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time in cognitive
psychological research, the logarithm
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

To better understand this model, we can identify the gamma model in terms of the
commonly used shape and scale parameters. For a gamma distribution with
shape~$a$ and scale~$b$, we have mean $\mu = ab$ and variance $\nu = ab^2$. The
variance as a function of the mean~$\mu$ in the GLM sense is
$\text{var}(\mu) = \mu^2 / a$, which implies that the dispersion parameter is
$\phi = 1 / a$. Therefore, for the gamma mixed-effect model, the conditional
gamma distribution of the reaction time $Y_{ijk}$ given the venue
effect~$v_i$ and heat effect~$h_{i/j}$
has shape $a = 1 / \phi$ and scale
\jy{Note that, as used in the code, we should have
  \texttt{sigma(model) \^ 2 =  phi}. So in Table 2, we can report phi in the last column.}
$b = g^{-1}(\alpha + v_i + h_{i/j}) \phi$. The marginal
distribution of $Y_{ijk}$ is a scale-mixture of gamma distributions, which can be
easily simulated from once the parameters are estimated. Many
random numbers generated from the fitted mixture distribution can be used to
approximate the probability of observing a reaction time faster than any given
threshold.  We are specifically interested in the probability of a reaction time
being less than 0.1 seconds in order to gauge if that is a reasonable 
disqualification barrier.

@@

@@

Table~\ref{tab:Clusrankresults} shows the results generated from both the 
permutation (exact) and asymptotic rank-based tests for the three comparisons, 
as discussed previously in Section~\ref{sec:rank}.    
The national versus international comparison
produced lower p-values than the 2019 versus 2022 comparison, with the
national vs international permutation test producing a p-value of 
$2.2\cdot10^{-16}$, which is the smallest possible value achievable in R.  
\eds{Prof. Yan, is this possible based on 1000 permutations?}
\jy{That cannot be true.  Redo with a much bigger number of permutations and
  update the table.}
This
is a highly significant test result and shows that athletes who competed at the
2022 World Track and Field Championships were on average, significantly faster
at reacting than they were just a few months prior.  For the 2019 versus 2022
same athlete comparison with permutation test, the p-value was 
$2.9\cdot10^{-3}$, which is significant at $\alpha = 0.01$.  The result of
comparing 2022 and 2023 was very similar to the above result, as the p-value
from the permutation test was $3.0\cdot10^{-3}$.  Taken together, these three
tests help show that athletes competing at the 2022 World Track and Field
Championships had faster reaction times than when they had competed previously.

@@

This paper does not consider data from the women's 100 meter hurdle, and although
that was initially because of concerns over reaction time differences in men and
women, exploratory analysis showed that 2022 was at best only
slightly faster relative to other championships going back to 2003.
Likewise, the rank-based comparisons resulted in p-values
that were not drastically significant after adjusting for multiple tests.
The lack of evidence from the women's data makes it 
harder to conclude that the sensors used at the 2022 championships were faulty, 
although it is unknown if the sensors for the men's 110 meter hurdle and women's
100 meter hurdle were the same.  On the other hand, this may be related to 
another debated question regarding whether the reaction time barrier should be 
the same for men and women.

It is confusing as to why the reaction time barrier for men and women 
is the same despite numerous articles \citep[e.g.,][]{lipps2011implications, 
babicc2009reaction, panoutsakopoulos2020gender} stating differently.  
According to these authors, World Athletics is ignoring differences in gender by 
setting the reaction time barrier to be equal. 
If there are gender differences and 0.1 seconds is considered a fair time for 
women, then 0.1 seconds could not be considered fair for men, as the probability 
of observing reaction times lower than 0.1 would be elevated.
\citet{brosnan2017effects} advocated for gender based reaction time barriers 
which seems logical if there are biological
differences in reaction times between men and women.

The first part of this paper showed that athletes' reaction times at the 2022 
World Track and Field Championships were generally faster than their reaction 
times at other competitions. Table~\ref{tab:Clusrankresults} shows multiple 
significant p-values that indicate differences in average reaction times for 
athletes who competed at the 2022 World Track and Field Championship and another
competition.  Additionally, this paper has shown that a reaction time of 0.1 
seconds is not as extreme as one may think, and that based on the probabilities
in Table~\ref{tab:Sim_time}, World Athletics may consider changing the 
disqualification barrier to 0.8 seconds.  If they were to do so, they would 
allow athletes like Allen to be able to react faster without fear of being 
disqualified.  Taken together, the results show that 2022 was an outlier and 
despite being disqualified, Allen's time was not so extreme.
